Determinism, Complexity, and Predictability in Computer Performance

### 2014-07-22 response.tex

Computer dynamics spans the full range of behavior reported in the
nonlinear dynamics literature.  Standard nonlinear time-series
analysis of data from these systems has confirmed the existence of
fixed points, periodic orbits of different periods, chaotic
trajectories with various Lyapunov exponents, etc.  in their dynamics.
While it is certainly possible that the nonlinear dynamics of computer
performance is subtly different from the nonlinear dynamics of other
systems---in a manner that cannot be detected by traditional nonlinear
time-series analysis---we believe that our results will hold for time
series drawn from other physical systems.

We are certainly aware that this claim may not be true, however, and
we are in the process of the follow-up study suggested by the
reviewer.  We have found that the relationship posited in our paper
holds for a time series of voltages from a chaotic laser used in the
SFI prediction competition, for instance.  (It also holds for various
paradigmatic systems, such as the Henon map and the Lorenz system, as
discussed in a later section of this response.)  The image below is an
augmented version of our revised Figure 7 that includes all of these
examples.

\begin{center}
    \includegraphics[width=0.8\columnwidth]{figs/new_prediction_vs_entropy_extras}
\end{center}

\noindent The dashed curve in this plot (and in Figure 6) was fit
\emph{solely on the computer performance data}.  Nonetheless, the
points from the other time series fall within a narrow band around the
fit.  Note that this fit ($y = a \log(b x + 1) + c x^16$) is different than in
the submitted version.  A careful examination of the data, prompted by
the reviewer's questions, suggested that this was a better match to
the data; theoretically, too, a log fit is more appropriate because
any signal with zero entropy should be perfectly predictable, and so
should ideally have a MASE value of zero.  Finally, note that there
are a lot more points on this plot than in our original version; this
is discussed further below.

@@

We have added a short discussion about the differences in notions and
definitions of complexity to the introduction and related-work
section, and clarified the working definition that we use in this
paper.

\smallskip

\emph{4) The second finding reads ``The way information is generated
  and processed internally by a system plays a crucial role in the
  success of different forecasting schema - and in the choice of which
  one is appropriate for a given time series''. This statement also
  looks trivial. Indeed, an optimal predictor for a linear stochastic
  system and low-dimensional nonlinear deterministic system are quite
  different (an AR model versus a local or global nonlinear model) due
  to different properties of the original systems. Thus, there is
  nothing to be proven here. Probably, the authors imply their
  particular benchmark signals (reflecting a computer behavior) but
  they do not claim that.  Thus, both primary findings seem to provide
  nothing new.}

The reviewer is absolutely right.  This finding, as it was originally
stated, is indeed trivial.  Our intent, as we have modified the text
to reflect, is that the appropriateness of a predictor can be
discovered \emph{empirically}, via comparison of its MASE value with
its WPE value.  Previous explorations of predictability have used
clean data and/or systems where the generating process is known.  Our
work makes neither of these assumptions.

Given a linear stochastic system, a linear stochastic prediction model
should certainly be more productive than a nonlinear deterministic
prediction model.  But in real-world data, the generating process is
almost never known, and it can be very hard to evaluate a model that
one has constructed.  Standard methods like the AIC approach used in
ARIMA modeling make a choice from among a large class of possible
models.  These selection algorithms rarely have a ``none of these are
appropriate'' output; they simply choose the model that best satisfies
their built-in criteria---even if none of the models in the class are
really appropriate.

The challenge for the practitioner is to know whether there is a
mismatch between a particular time series and a particular forecast
model: i.e., whether one can do better and should try another model,
or whether the time series is so complicated that the model is doing
as well as one could expect.  The WPE vs. MASE curve in our paper is
an empirical heuristic that should help practitioners do that.

We appreciate the reviewer alerting us to our muddy wording and
thinking about these issues.  We have reworked the findings to clarify
matters and also added an explanatory paragraph directly following
those findings.  While this lengthened the introduction, we feel that
it has focused and clarified the contribution of the work.

@@

\emph{II) PE is invariant under an invertible nonlinear change of
  variables. The prediction error measured as the authors suggest is
  not invariant. Thus, such a simple change of variables can lead, at
  least, to a change in regression coefficient between WPE and
  prediction error. Thus, the authors should also discuss to which
  variable for that class of systems their result apply. I think that
  the fact that the authors observe a certain regression line at all
  seems to be due to small number of different signals they consider
  (only 7 or 8 different data points, such a small number of points
  may lie approximately near a straight line by chance).}

\alert{Liz to fix following text, now that we have these experiments}

PE is invariant under an invertible nonlinear transformation of the
values in the time series, but WPE is not.  But MASE varies \emph{in a
  complementary way}.  Ryan will do $x(t)$ and $5e^{x(t)}$ and show
that it's on the curve.

% Ryan will insert figure here with x(t) = svdone and svdsix
\begin{center}
    \includegraphics[width=0.8\columnwidth]{figs/nonlinear_transform_data}
\end{center}

@@

\alert{Need the change and the confidence stuff for this: Fit changed
  by [[what]] when we redid it with these 34 new points.}  It now
reflects a map, a flow, a purely stochastic signal, a variety of
real-world data from two different systems (laser, computer), and
nonlinear transformations of that real-world data.  Seems pretty
compelling.

\emph{To summarize, the paper shows that the authors are well familiar
  with the entire topic and highly qualified, they overview some
  well-known facts (WPE, various prediction techniques, etc) and
  ``invent'' an unexpected test system (computer performance data),
  they report about their attempts to play around with all that
  staff. However, their results do not seem to provide any new
  information to a specialist. Probably, this work can be continued to
  get some more reliable results about concrete relationship between
  WPE and prediction error for certain classes of systems (however, I
  doubt that it is possible to derive any simple and widely applicable
  relationship like the linear function suggested by the
  authors). Seemingly, it should then be a new paper which can be
  hardly considered as a revised version of the present manuscript.}

It is our hope that these clarifications, the attendant modifications
to the paper, and the additional preliminary results that we included
in this rebuttal letter have convinced the reviewer that this
manuscript is appropriate for publication in PRE.

Because of the depth and extent of the revisions that we made in
response to this reviewer's comments, we did not feel that it was
effective to highlight the changes in a different color in the
manuscript; please let us know if a more-detailed description of the
changes would make the re-reading easier.

### 2014-07-22 intro.tex
Liz editing doc and response wrt findings

This leads naturally to an important and challenging question: given a
noisy real-valued empirical time series, does there exist a forecast
model that can leverage the information (if any) that is being
transmitted from past to future by the underlying generating process?
A first step in answering this question is to reliably quantify where
on the complexity spectrum a given time series falls; a second step is
to determine how complexity and predictability are related in these
kinds of data sets.  With these answers in hand, one can develop a
practical strategy for assessing appropriateness of forecast methods
for a given time series.  If the forecast produced by a particular
method is poor, for instance, but the time series contains a
significant amount of predictive structure, one can reasonably
conclude that that method is inadequate to the task and that one
should seek another method.  The goal of this paper is to explore and
formalize that strategy.

@@

We focus on real-valued, scalar, time-series data.
%
% that appears complicated: i.e., not periodic, linear, constant, etc.
%
We do not make any knowledge of the generating process or its
properties: whether it is linear, nonlinear, deterministic,
stochastic, stationary, non-stationary, etc.  To explore the
relationship between complexity, predictive structure, and
predictability, we generate forecasts for a variety of time-series
datasets using four different prediction methods, then compare the
accuracy of those predictions to the permutation entropy of the
associated signals.  This results in two primary findings:
\begin{enumerate}
\item The complexity of a noisy real-valued time series is \emph{(i)}
  quantifiable by permutation entropy and \emph{(ii)} correlated with
  the accuracy of an appropriate predictor.
\item The way information is generated and processed internally by a
  system plays a crucial role in the success of different forecasting
  schema---and in the choice of which one is appropriate for a given
  empirical time series.
\end{enumerate}
There has, of course, been a great deal of good work on different ways
to measure the complexity of data, and previous explorations have
confirmed repeatedly that complexity is a challenge to prediction, but
this collection of issues has not been properly explored in the
context of noisy, poorly sampled, real-world data from unknown
systems.  That is the contribution of this paper.

@@

For the purposes of this study, we require a broad array of
time-series datasets from across the complexity spectrum.  We chose to
study sensor data from a computer-performance experiment.  While this
is not a common laboratory experiment, it is a highly appropriate
choice here.  Computers are extremely complicated systems and their
dynamics is surprisingly rich.
% Modern microprocessor chips contain multiple processing units and
% multi-layer memories, for instance, and they use complicated
% hardware/software strategies to maximize performance by moving data
% and threads of computation across those resources.  As a result, the
The processor and memory loads during the execution of even a very
simple program can exhibit dynamical chaos, for
instance~\cite{mytkowicz09}.  Figure~\ref{fig:col-ipc} shows an
example: a performance trace of a short program that repeatedly
initializes the upper triangle of a matrix in column-major order.
%
 \begin{figure}[htp]
    \centering
    \includegraphics[width=\columnwidth]{figs/colshortts}
    % where an .eps filename suffix will be assumed under latex,
    % and a .pdf suffix will be assumed for pdflatex
    \caption{A computer performance trace: the processor load of \col,
      a simple C program that repeatedly initializes a matrix in
      column-major order, running on an Intel
      i7\textsuperscript{\textregistered}-based machine.}
   \label{fig:col-ipc}
  \end{figure}
%
A small change in the code can cause the dynamics to bifurcate to a
periodic regime.  
% 
% Indeed, many computer performance traces exhibit
% interesting regime changes as a program moves through the different
% phases of its operation.
By running different programs on the same computer, then, we can
produce traces that span the whole range of the complexity spectrum,
from completely predictable to completely unstructured---which makes
this an ideal testbed for this study\footnote{Predicting the
  \emph{state} of a computer, of course, would amount to solving the
  halting problem.  What we are doing here is predicting computer
  \emph{performance}, which does not violate the Rice-Shapiro
  theorem~\cite{hopcroft2007}.}.
% footnote to appease the theoretician in Josh

>>>>

This leads naturally to an important and challenging question: given a
noisy real-valued empirical time series, does there exist a forecast
model that can leverage the information (if any) that is being
transmitted from past to future by the underlying generating process?
A first step in answering this question is to reliably quantify where
on the complexity spectrum a given time series falls; a second step is
to determine how complexity and predictability are related in these
kinds of data sets.  With these answers in hand, one can develop a
practical strategy for assessing appropriateness of forecast methods
for a given time series.  If the forecast produced by a particular
method is poor, for instance, but the time series contains a
significant amount of predictive structure, one can reasonably
conclude that that method is inadequate to the task and that one
should seek another method.  The goal of this paper is to develop
effective, practical heuristics for that strategy.

@@

We focus on real-valued, scalar, time-series data from physical experiments.
%
% that appears complicated: i.e., not periodic, linear, constant, etc.
%
We do not make any knowledge of the generating process or its
properties: whether it is linear, nonlinear, deterministic,
stochastic, stationary, non-stationary, etc.  To explore the
relationship between complexity, predictive structure, and
predictability, we generate forecasts for a variety of experimental
time-series datasets using four different prediction methods, then
compare the accuracy of those predictions to the permutation entropy
of the associated signals.  This results in two primary findings:
% 
\begin{enumerate}
% \item The complexity of a noisy real-valued time series can be
%   effectively quantifiable by computing its permutation entropy.

\item The permutation entropy of a noisy real-valued time series is
  correlated with the accuracy of an appropriate predictor.

\item The relationship between permutation entropy and prediction
  accuracy is a useful empirical heuristic for identifying when there
  is a mismatch between a time series from an unknown system and a
  given prediction model.
\end{enumerate}
% 
There has, of course, been a great deal of good work on different ways
to measure the complexity of data, and previous explorations have
confirmed repeatedly that complexity is a challenge to prediction.  It
is well known that the way information is generated and processed
internally by a system plays a crucial role in the success of
different forecasting methods---and in the choice of which method is
appropriate for a given time series.  
This constellation of issues has not been properly explored, however,
in the context of noisy, poorly sampled, real-world data from unknown
systems.  That exploration---and the development of strategies for
putting its results into effective practice---is the primary
contribution of this work.  The empirical results in the later
sections of this paper not only elucidate the relationship between
complexity and predictability.  The strategy that we derive from these
results can also aid practitioners in choosing an appropriate
prediction model for a given real-world noisy time series from an
unknown system---a challenging task for which little guidance is
currently available.

\alert{With all the extra material above, the following paragraph now
  feels repetitive.  Given that, and the reviewer's comment that the
  intro is too long, I suggest that we merge it into a later section:}

@@

\alert{This paragraph, too, might well be moved to the beginning of
  section 3:}

For the purposes of this study, we required a broad array of
time-series datasets from across the complexity spectrum.  We chose to
study sensor data from a computer-performance experiment.  While this
is not a common laboratory experiment, it is a highly appropriate
choice here.  Computers are extremely complicated systems and their
dynamics is surprisingly rich.
% Modern microprocessor chips contain multiple processing units and
% multi-layer memories, for instance, and they use complicated
% hardware/software strategies to maximize performance by moving data
% and threads of computation across those resources.  As a result, the
The processor and memory loads during the execution of even a very
simple program can exhibit dynamical chaos, for
instance~\cite{mytkowicz09}.  Figure~\ref{fig:col-ipc} shows an
example: a performance trace of a short program that repeatedly
initializes the upper triangle of a matrix in column-major order.
%
 \begin{figure}[htp]
    \centering
    \includegraphics[width=\columnwidth]{figs/colshortts}
    % where an .eps filename suffix will be assumed under latex,
    % and a .pdf suffix will be assumed for pdflatex
    \caption{A computer performance trace: the processor load of \col,
      a simple C program that repeatedly initializes a matrix in
      column-major order, running on an Intel
      i7\textsuperscript{\textregistered}-based machine.}
   \label{fig:col-ipc}
  \end{figure}
%
A small change in the code can cause the dynamics to bifurcate to a
periodic regime.  
% 
% Indeed, many computer performance traces exhibit
% interesting regime changes as a program moves through the different
% phases of its operation.
By running different programs on the same computer, then, we can
produce traces that span the whole range of the complexity spectrum,
from completely predictable to completely unstructured---which makes
this an ideal testbed for this study\footnote{Predicting the
  \emph{state} of a computer, of course, would amount to solving the
  halting problem.  What we are doing here is predicting computer
  \emph{performance}, which does not violate the Rice-Shapiro
  theorem~\cite{hopcroft2007}.}.
% footnote to appease the theoretician in Josh

### 2014-07-22 related.tex
Liz editing doc and response wrt findings

The model-based error analysis methods described in the previous three
paragraphs are based on specific assumptions about the underlying
generating process and knowledge about what will happen to the error
if those assumptions hold or fail.  Model-free information analysis
moves away from those restrictions.  Our approach falls into this
class.  Our goal is to measure the inherent complexity of an empirical
time series, then correlate that complexity with the predictive
accuracy of forecasts made using a number of different methods.

@@

The notion of complexity used here differs from that considered by
e.g. \cite{Shalizi2008}, which would consider a time series without
any statistical regularities to be non-complex.

@@

As a final note: there has been prior work under a similar title to
ours~\cite{haven2005}, but there are only superficial similarities
between the two research projects.  Haven \emph{et al.} utilize the
relative entropy to quantify the difference in predictability between
two distributions: one evolved from a small ensemble of past states
using the known dynamical system, and the other the observed
distribution. Our work quantifies the predictability of a single
observed time series using weighted permutation entropy and makes no
assumptions about the generating process.  More closely related is the
work by Boffetta et al.~\cite{boffetta02}.  \alert{[[Sentences from
      Ryan explaining exactly what the difference is.]]]

>>>>

The model-based error analysis methods described in the previous three
paragraphs are based on specific assumptions about the underlying
generating process and knowledge about what will happen to the error
if those assumptions hold or fail.  Model-free information analysis
moves away from those restrictions.  Our approach falls into this
class: we wish to measure the inherent complexity of an empirical time
series, then correlate that complexity with the predictive accuracy of
forecasts made using a number of different methods.

@@

@@

There has been prior work under a similar title to
ours~\cite{haven2005}, but there are only superficial similarities
between the two research projects. Haven \emph{et al.} utilize the
relative entropy to quantify the difference in predictability between
two distributions: one evolved from a small ensemble of past states
using the known dynamical system, and the other the observed
distribution. Our work quantifies the predictability of a single
observed time series using weighted permutation entropy and makes no
assumptions about the generating process.

More closely related is the work by Boffetta \emph{et
  al.}~\cite{boffetta02}, who investigate the scaling behavior of
finite-time Lyapunov exponents (FTLE) and $\epsilon$-entropy for a
wide variety of deterministic systems with known dynamics.  While the
scaling of these measures acts as a general proxy for predictability
bounds, this approach differs from our work in a number of fundamental
ways.  First, we focus on real-world time-series data, where one does
not necessarily have the ability to perturb or otherwise interact with
the system of interest, nor can one obtain or manufacture the
(possibly large) number of points that might be needed to estimate the
$\epsilon$-entropy for small $\epsilon$.  Second, we tie
information---in the form of the weighted permutation
entropy---directly to prediction error via MASE values.  Though FTLE
and $\epsilon$-entropy allow for the comparison of predictability
between systems, they do not directly provide an estimate of
prediction error.  [[Finally, our approach also holds for
    \emph{stochastic} systems, where the computation of FTLEs makes no
    sense and/or doesn't reflect predictability.  I remember someone
    saying this but didn't see it here, so I added it.  Boffetta et
    al. just do dynamical systems, yes?  And lambda of a stochastic
    system wouldnt' make sense, yes?]]
